---
title: "***zoolog***:  \n Zooarcheological Analysis with Log-Ratios"
author: "Jose M Pozo, Silvia Valenzuela-Lamas, Angela Trentacoste,  Ariadna Nieto-Espinet and Silvia Guimarães Chiarelli"
email: "josmpozo@gmail.com, svalenzuela@imf.csic.es"
date: "`r format(Sys.Date())`"
bibliography: ../inst/REFERENCES.bib
## to create the vignettes 'outside' the package, with table of content (toc)
#output:
#  html_document:
#    toc: true
#    toc_float:
#      collapsed: false
#      smooth_scroll: false
## to create the vignettes 'inside' the package, without table of content (toc)
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zoolog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The package *zoolog* includes functions and reference data to generate and 
manipulate log-ratios (also known as log size index (LSI) values) from measurements obtained on zooarchaeological material. Log ratios are used to compare the relative (rather than the absolute) dimensions of animals from archaeological contexts [@meadow1999use]. Essentially, the method compares archaeological measurements to a standard, producing a value that indicates how much larger or smaller the archaeological specimen is compared to that standard. *zoolog* is also able to seamlessly integrate data and references with heterogeneous nomenclature, which is internally managed by a zoolog thesaurus.

The methods included in the package were first developed in the framework of the 
ERC-Starting Grant 716298 ZooMWest (PI S. Valenzuela-Lamas), and were first used 
in the paper [@trentacoste2018pre]. They are based on the techniques proposed by @simpson1941large and @simpson1960quantitative, which calculates *log size index* (LSI) values as: 
$$
\mbox{LSI} = \log x - \log x_{\text{ref}} = \log(x/x_\text{ref}),
$$
where $x$ is the considered measure value and $x_\text{ref}$ is the corresponding reference value.

Several different sets of standard reference values are included in the package. These standards include several published and widely used biometric datasets (e.g. @davis1996measurements, @albarella2005neolithic) as well as other less known standards. These references, as well as the data example provided with the package, are based on the measures and measure abbreviations defined in @von1976guide and @davis1992rapid. 

In general, zooarcheological datasets are composed of skeletal remains representing many different anatomical body parts. In investigation of animal size, the analysis of measurements from a given anatomical element provides the best control for the variables affecting size and shape and, as such, it is the preferable option. Unfortunately, this approach is not always viable due to low sample sizes in some archaeological assemblages. This problem can be mitigated by calculating the LSI values for measurements with respect to a reference, which provides a means of aggregating biometric information from different body parts. The resulting log ratios can be compared and statistically analysed under reasonable conditions [@albarella2002size]. However, length and width measurements of different anatomical elements still should not be directly compared or aggregated for statistical analysis. 

The package includes a *zoolog thesaurus* to facilitate its usage by research teams across the globe, and working in different languages and with different recording traditions. The thesaurus enables the *zoolog* package to recognises many different names for taxa and skeletal elements (e.g. "Bos taurus", "Bos", "BT",  "bovino", "bota"). Consequently, there is no need to use a particular, standardised recording code for the names of different taxa or elements. 

## Acknowledgements
We are particularly grateful to Sabine Deschler-Erb and Barbara Stopp, from the University of Basel (Switzerland) for making the reference values of several specimens available through the ICAZ Roman Period Working Group, which have been included here with their permission. We also thank Francesca Slim and Dimitris Filioglou from the University of Groningen for providing additional reference sets.

The thesaurus has benefited from the contributions from Moussab Besso, Canan Cakirlar, Jwana Chahoud, Jacopo De Grossi Mazzorin, Sabine Deschler-Erb, Dimitrios Filioglou, Armelle Gardeisen, Sierra Harding, Pilar Iborra, Michael MacKinnon, Nimrod Marom, Claudia Minniti, Francesca Slim, Barbara Stopp, and Emmanuelle Vila.

We are grateful to all of them for their contributions, comments, and help. In addition, users are encouraged to contribute to the thesaurus and other references so that *zoolog* can be expanded and adapted to any database.


# Installation

You can install the released version of zoolog from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("zoolog")
```

And the development version from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("josempozo/zoolog@HEAD", build_vignettes = TRUE, force = TRUE)
```


# Reference

The package *zoolog* includes several osteometrical references. Currently, the references include reference values for the main domesticates and their agriotypes (*Bos*, *Ovis*, *Capra*, *Sus*), and reed deer (*Cervus elaphus*), originated from a variety of publications and resources. In addition, the user can consider other references, or the provided references can be extended and updated integrating newer research data. Submission of extended/improved references is encouraged. Please, contact the maintainer through the provided email address to make the new reference fully accessible.

The predefined reference sets are provided in the named list `reference`, which currently include the following `r length(zoolog::reference)` sets:
```{r}
library(zoolog)
str(reference, max.level = 1)
```
The reference set `reference$Combi` is the default reference for computing the [log ratios](../help/LogRatios), since it includes the most comprehensive reference for each species.

This reference sets are build by selecting cases from the `referencesDatabase` formed by the taxon-specific references from all the considered resources. The selection is defined by the data frame `referenceSets`:
```{r echo=FALSE}
options(knitr.kable.NA = "")
knitr::kable(referenceSets)
```

A detailed description of the reference data, including its structure, properties, and considered resources can be found in the [ReferencesDatabase](../help/referencesDatabase) help page.

# Thesaurus

A [thesaurus set](../help/zoologThesaurus) is defined in order to make the methods robust to different nomenclatures used in datasets created by different authors, with different conventions and in different languages. This enables the function [LogRatios](../help/LogRatios) to recognize the relevant names in the reference that match each data entry regardless of the used nomenclature, as long as it is included in the relevant thesaurus. The thesaurus also allows to [standardize the nomenclature](../help/StandardizeNomenclature) of the dataset if desired. 

The user can also use other thesaurus sets or modify the provided one. In this latter case, we encourage the user to contact the maintainer at the provided email address so that the additions can be incorporated into the new versions of the package.

Currently, the zoolog thesaurus set includes four thesauri:

identifierThesaurus
  : It refers to the column names identifying the relevant variables to match
    the data with the corresponding reference in computing the log ratios. It
    includes the categories *Taxon*, *Element*, *Measure*, and 
    *Standard*, each providing different equivalent names. For instance,
    *Taxon* includes the options 
    `r a <- zoologThesaurus$identifier$Taxon; 
      paste0("*", paste(a[a != ""], collapse = "*, *"), "*")`.

taxonThesaurus
  : Naming of the different taxa when recording animal
    bones. The current categories are 
    *`r paste(names(zoologThesaurus$taxon), collapse = "*, *")`*,
    each providing different equivalent names. This thesaurus is case, accent, and     punctuation insensitive, so that, for instance, "Bos" is equivalent to 
    "bos", "Bos.", or "Bos\ ".

elementThesaurus
  : Naming of anatomical elements when recording animal bones. It currently
    includes `r ncol(zoologThesaurus$element)` categories
    (*`r paste(names(zoologThesaurus$element)[1:3], collapse = "*, *")`*, ...),
    each providing different equivalent names. 
    It is case, accent, and punctuation insensitive.


measureThesaurus
  : Naming of the measurements. In general, 
    the English abbreviations from [@von1976guide] and [@davis1992rapid] are
    widely used. But this enables other nomenclatures (e.g. original German
    abbreviations in [@von1976guide]) to be included. This thesaurus is case
    sensitive.

# Functions
The full list of functions is available under the zoolog *help* page. We list them here sorted by its prominence for a typical user, and grouped by functionality:

[**LogRatios**](../help/LogRatios)
  : It computes the log ratios of the measurements in a dataset relative to 
    standard reference values. By default 'Reference$Combi' is used. The 
    function includes the option 'joinCategories' allowing several taxa 
    (typically *Ovis*, *Capra*, and unknown *Ovis/Capra*) to be considered
    together with the same reference taxon.

[**CondenseLogs**](../help/CondenseLogs)
  : It condenses the calculated log ratio values into a reduced number of 
    features by grouping several measure log ratios and selecting or calculating 
    a representative feature value. By default the selected groups represent a
    single dimension, i.e. *Length* and *Width*. Only one feature is extracted 
    per group. Currently, two methods are possible: "priority" (default) or
    "average". 
  : This operation is motivated by two circumstances. First, not all
    measurements are available for every bone specimen, which obstructs their
    direct comparison and statistical analysis. Second, several measurements 
    can be strongly correlated (e.g. SD and Bd both represent bone width). 
    Thus, considering them as independent would produce an over-representation 
    of bone remains with multiple measurements per axis. Condensing each group 
    of measurements into a single feature (e.g. one measure per axis) alleviates
    both problems. 
  : The default method ("priority"), selects the first available log ratio in 
    each group. We provide the following by-default group and prioritization: 
    For lengths, the first option is GL, then HTC. For widths, the order of
    priority is: Bd, BT, Bp, SD. This order maximises the robustness and
    reliability of the measurements, as priority is given to the most abundant,
    more replicable, and less age dependent measurements.The method 'average'
    extracts the mean per group, ignoring the non-available log ratios.

[**RemoveNACases**](../help/RemoveNACases)
  : It removes the cases (table rows) for which all measurements of interest are
    non-available (NA). A particular list of measurement names can be explicitly
    provided or selected by a common initial pattern (e.g. prefix). The default
    setting removes the rows with no available log ratios to facilitate 
    subsequent analysis of the data.

[**InCategory**](../help/InCategory)
  : It checks if an element belongs to a category according to a thesarus. 
    It is similar to [base::is.element](../help/is.element), returning a logical
    vector indicating if each element in a given vector is included in a given
    set. 
    But `InCategory` check for equality assuming the equivalencies defined in 
    the given thesaurus. It is intended for the user to easily select a subset
    of data without having to standardize the analysed dataset.

[**Nomenclature standardization**](../help/StandardizeNomenclature)
  : This includes two functions enabling the user to map data with heterogeneous
    nomenclature into a standard one as defined in a thesaurus:
    
    * `StandardizeNomenclature` standardizes a character vector according to 
      a given thesaurus.
    * `StandardizeDataSet` standardizes column names and values of a data 
      frame according to a thesaurus set.

[**AssembleReference**](../help/AssembleReference)
  : It allows the user to build new references assembling the desired
    taxon-specific references included in the *zoolog* `referencesDataSet` or
    in any other provided by the user.

[**Thesaurus readers and writers**](../help/ThesaurusReaderWriter)
  : This includes functions to read and write a single thesaurus 
    (`ReadThesaurus` and `WriteThesaurus`) and a thesaurus set
    (`ReadThesaurusSet` and `WriteThesaurusSet`).
    
[**Thesaurus management**](../help/ThesaurusManagement)
  : This includes functions to modify and check thesauri:
  
    * `NewThesaurus` generates an empty thesarus.
    * `AddToThesaurus` adds new names and categories to an existing thesaurus.
    * `RemoveRepeatedNames` cleans a thesarus from any repeated names on any 
      category.
    * `ThesaurusAmbiguity` checks if there are names included in more than one
      category in a thesaurus.


# Examples
The following examples are designed to be read and run sequentially. They 
represent a possible pipeline, meaningful for the processing and analysis of a
dataset. Only occasionally, a small diversion is included to illustrate some alternatives.

## Reading data and calculating log ratios
This example reads a dataset from a file in csv format and computes the
log-ratios. Then, the cases with no available log-ratios are removed.
Finally, the resulting dataset is saved in a file in csv format.

The first step is to set the local path to the folder where you have the dataset to be analysed (this is typically a comma-separated value (csv) file). Here the [example dataset](../help/dataValenzuelaLamas2008) from [@valenzuela2008alimentacio] included in the package is used:

```{r}
library(zoolog)
dataFile <- system.file("extdata", "dataValenzuelaLamas2008.csv.gz", 
                        package = "zoolog")
data = read.csv2(dataFile,
                 quote = "\"", header = TRUE, na.strings = "",
                 fileEncoding = "UTF-8",
                 stringsAsFactors = TRUE)
knitr::kable(head(data)[, -c(6:20,32:64)])
```
To enhance the visibility, we have shown only the most relevant columns.

We now calculate the log-ratios using the function `LogRatios`. Only measurements that have an associated standard will be included in this calculation. The log values will appear as new columns with the prefix 'log' following the original columns with the raw measurements:
```{r}
dataWithLog <- LogRatios(data)
knitr::kable(head(dataWithLog)[, -c(6:20,32:64)])
```

## Dealing with *lazy* datasets
If we observe more carefully the example dataset, we can see that the *astragali* have been assigned a *GL* measurement, instead of *GLl*. For this we can use the function [InCategory](../help/InCategory), which allows to select cases taking into account the thesaurus:
```{r}
AScases <- InCategory(dataWithLog$Os, "talus", zoologThesaurus$element)
knitr::kable(head(dataWithLog[AScases, -c(6:20,32:64)]))
```
This is a dataset simplification that has been used by some researchers. It is possible because *GLl* is only relevant for the astragalus, while *GL* is not applicable to it. Thus, there cannot be any ambiguity between both measures since they can be identified by the bone element. However, since the reference uses the proper measure name for each bone element (*GLl* for the astragalus), the reference measure has not been correctly identified. Consequently, the log ratio *logGL* has `NA` values and the column *logGLl* does not exists.

The optional parameter mergedMeasures facilitates the processing of this type of simplified datasets. For the example data, we can use
```{r}
GLandGLl <- list(c("GL", "GLl"))
dataWithLog <- LogRatios(data, mergedMeasures = GLandGLl)
knitr::kable(head(dataWithLog[AScases, -c(6:20,32:64)]))
```
This option allows to automatically select, for each bone element, the corresponding measure present in the reference. Observe that now the log ratios have been computed and assigned to the column *logGL*.

## Using the same ovis reference for all caprines
We could be interested in obtaining the log ratios of all caprines, including *Ovis aries*, *Capra hircus*, and undetermined *Ovis/Capra*, with respect to the reference for *Ovis aries*. This can be set using the argument `joinCategories`.
```{r}
caprineCategory <- list(ovar = c("sheep", "capra", "oc"))
dataWithLog=LogRatios(data, joinCategories = caprineCategory, mergedMeasures = GLandGLl)
knitr::kable(head(dataWithLog)[, -c(6:20,32:64)])
```
Note that this option does not remove the distinction in the data between the different species, it just indicates that for these taxa the log ratios must be computed from the same reference (`ovar`).

## Pruning the data from cases with no available measure
The cases without log-ratios can be removed to facilitate subsequent analyses:
```{r}
dataWithLogPruned=RemoveNACases(dataWithLog)
knitr::kable(head(dataWithLogPruned[, -c(6:20,32:64)]))
```

You may want to write the resulting file in the working directory (you need to set it first):
```{r, eval = FALSE}
write.csv2(dataWithLogPruned, "myDataWithLogValues.csv", 
           quote=FALSE, row.names=FALSE, na="", 
           fileEncoding="UTF-8")
```

## Condensing log values
After calculating log ratios using the `LogRatios` function, many rows in the resultant dataframe (dataWithLog in the example above) may contain multiple log values, i.e. you will have several log values associated with a particular archaeological specimen. When analysing log ratios, it is preferential to avoid overrepresentation of bones with a greater number of measurements and account for each specimen only once. The `CondenseLogs` function extracts one length and one width value from each row and places these in new Length and Width columns.The 'priority' method described in [@trentacoste2018pre] has been set as default. Nevertheless, other options (e.g. average of all width log values for a given specimen) can be chosen if preferred. In this case, the default option has been used:
```{r}
dataWithSummary <- CondenseLogs(dataWithLogPruned)
knitr::kable(head(dataWithSummary)[, -c(6:20,32:64,72:86)])
```



## Standardizing the dataset nomenclature
The integration of the thesaurus functionality facilitates the use of datasets with heterogenous nomenclatures, without further preprocessing. An extensive catalogue of names for equivalent categories have been integrated in the provided thesaurus set `zoologThesaurus`. These equivalences are internally and silently managed without requiring any action from the user. However, it can be also interesting to explicitly standardize the data to make figures legible to a wider audience. This is especially useful when different nomenclature for the same concept is found in the same dataset, for instance "sheep" and "ovis" for the same taxon or "hum" and "HU" for the bone element.

In the example data, `zoologThesaurus` will change for instance "ovar" to "Ovis aries".
```{r}
dataStandardized <- StandardizeDataSet(dataWithSummary)
knitr::kable(head(dataStandardized)[, -c(6:20,32:64,72:86)])
```

## Selecting only caprines
We may be interested in selecting all caprine elements. This can be done even without standardizing the data using the function `InCategory`:
```{r}
dataOC <- subset(dataWithSummary, InCategory(Especie, 
                                             c("sheep", "capra", "oc"),
                                             zoologThesaurus$taxon))
knitr::kable(head(dataOC)[, -c(6:20,32:64)])
```
Observe that no standardization is performed in the output subset. To standardize the subset data, `StandardizeDataSet` can be applied either before or after the subsetting.
```{r}
dataOCStandardized <- StandardizeDataSet(dataOC)
knitr::kable(head(dataOCStandardized)[, -c(6:20,32:64)])
```
Observe also that the distinction between *Ovis aries*, *Capra hircus*, and *Ovis/Capra* has not been removed from the data.

If we were interested only in one summary measure, *Width* or *Length*, we could  retain the cases including this measure:
```{r}
dataOCWithWidth <- RemoveNACases(dataOCStandardized, measureNames = "Width")
dataOCWithLength <- RemoveNACases(dataOCStandardized, measureNames = "Length")
```
which gives respectively `r nrow(dataOCWithWidth)` (`=nrow(dataOCWithWidth)`)  and `r nrow(dataOCWithLength)` (`=nrow(dataOCWithLength)`)  cases.

## Different plots for data visualisation
Condensed log values can be visualised as histograms and box plots using ggplot [@wickham2011ggplot2]. Here we will look at some examples of plotting values from caprines.

### Horizontal Boxplot with dots grouped by site

For the example plots we will use the package **ggplot2**.
```{r, echo = FALSE}
library(ggplot2)
```

We can now create a boxplot for the widths: 
```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
ggplot(dataOCStandardized, aes(x = Width, y = Site)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height = 0.2, width = 0, alpha = 1/2, color = 4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio")
```

And another boxplot for the lengths:
```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
ggplot(dataOCStandardized, aes(x = Length, y = Site)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height = 0.2, width = 0, alpha = 1/2, color = 4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine lengths") +
  xlab("Length log-ratio")
```

### Histograms grouped by site
We may choose to plot the width data as an histogram: 
```{r, fig.asp = 0.7, fig.width = 6, fig.align="center"}
ggplot(dataOCStandardized, aes(Width)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio") +
  facet_grid(Site ~.) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        axis.text = element_text(size = 10) ) +
  scale_y_continuous(breaks = c(0, 10, 20, 30))
```

### Vertical boxplot with dots grouped by taxon and site
Here we reorder the factor levels of `dataOCStandardized$Taxon` to make
the order of the boxplots more intuitive.

```{r}
levels0 <- levels(dataOCStandardized$Taxon)
levels0
```
```{r}
dataOCStandardized$Taxon <- factor(dataOCStandardized$Taxon, 
                                   levels = levels0[c(2,3,1)])
levels(dataOCStandardized$Taxon)
```
and assign specific colours for each category:
```{r, message = FALSE, fig.asp = 0.6, fig.width = 6, fig.align="center"}
Ocolour <- c("#A2A475", "#D8B70A", "#81A88D")
ggplot(dataOCStandardized, aes(x=Site, y=Width)) + 
  geom_boxplot(aes(fill=Taxon), 
               notch = TRUE, alpha = 0, lwd = 0.377, outlier.alpha = 0,
               width = 0.5, na.rm = TRUE,
               position = position_dodge(0.75),
               show.legend = FALSE) + 
  geom_point(aes(colour = Taxon, shape = Taxon), 
             alpha = 0.7, size = 0.8, 
             position = position_jitterdodge(jitter.width = 0.3),
             na.rm = TRUE) +
  scale_colour_manual(values=Ocolour) +
  scale_shape_manual(values=c(15, 18, 16)) +
  theme_bw(base_size = 8) +
  ylab("LSI value") +
  ggtitle("Sheep/goat LSI width values") 
```


### Histograms grouped by taxon and site  

```{r, warning = FALSE, fig.asp = 0.6, fig.width = 6, fig.align="center"}
TaxonSiteWidthHist <- ggplot(dataOCStandardized, aes(Width, fill = Taxon)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") + 
  ggtitle("Sheep/goat Widths") + facet_grid(Site ~ Taxon)
TaxonSiteWidthHist
```
```{r, warning = FALSE, fig.asp = 0.6, fig.width = 6, fig.align="center"}
TaxonSiteWidthHist <- ggplot(dataOCStandardized, aes(Width, fill = Taxon)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") + 
  ggtitle("Sheep/goat Widths") + facet_grid(~Site)
TaxonSiteWidthHist
```


## Statistical test
We may run a statistical test (here a Student t-test) to check whether the differences between sites have statistical significance on lengths:
```{r}
t.test(Length ~ Site, dataOCStandardized, 
       subset = Site %in% c("OLD", "ALP"))
```

or on widths:
```{r}
t.test(Width ~ Site, dataOCStandardized, 
       subset = Site %in% c("OLD", "ALP"))
```

For testing all possible pairs, the p-values must be adjusted for multiple comparisons:
```{r, message = FALSE}
library(stats)
pairwise.t.test(dataOCStandardized$Width, dataOCStandardized$Site, 
                pool.sd = FALSE)
```



# Bibliography
