---
title: "Package ***zoolog***:  \n Zooarcheological Analysis with Log-Ratios."
author: "Jose M Pozo, Silvia Valenzuela-Lamas, Angela Trentacoste,  Ariadna Nieto-Espinet and Silvia Guimarães Chiarelli"
email: "svalenzuela@imf.csic.es"
date: "`r format(Sys.Date())`"
bibliography: zoolog.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zoolog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The package ***zoolog*** includes functions and reference data to generate and 
manipulate log-ratios (also known as log size index (LSI) values) from measurements obtained on zooarchaeological material. Log ratios are used to compare the relative (rather than the absolute) dimensions of animals from archaeological contexts [@meadow1999use]. Essentially, the method compares archaeological measurements to a standard, producing a value that indicates how much larger or smaller the archaeological specimen is compared to that standard.

The methods included in the package were first developed in the framework of the 
ERC-Starting Grant 716298 ZooMWest (PI S. Valenzuela-Lamas), and were first used 
in the paper [@trentacoste2018pre]. They are based on the techniques proposed by [@simpson1941large] and [@simpson1960quantitative], which calculates LSI values as such: 

LSI (log size index) = (log x - log m) = log(x/m)


The zoolog package is based on the measurements and measurement abbreviations defined in [@von1976guide] and [@davis1992rapid]. Several different sets of standard reference values for calculating the log ratios are included in the package. These standards include several published and widely used biometric datasets (i.e. [@davis1996measurements], [@albarella2005neolithic]) as well as other less known standards. We are particularly grateful to Sabine Deschler-Erb and Barbara Stopp, from the University of Basel (Switzerland) for making the reference values of several specimens available through the ICAZ Roman Period Working Group, which have been included here with their permission. 

In general, zooarcheological datasets are composed of skeletal remains representing many different anatomical body parts. In investigation of animal size, the analysis of measurements from a given anatomical element provides the best control for the variables affecting size and shape and, as such, it is the preferable option. Unfortunately, this approach is not always viable due to low sample sizes in some archaeological assemblages. This problem can be mitigated by calculating the LSI values for measurements with respect to a reference, which provides a means of aggregating biometric information from different body parts. The resulting log ratios can be compared and statistically analysed under reasonable conditions [@albarella2002size]. However, length and width measurements of different anatomical elements still should not be directly compared or aggregated for statistical analysis. 

The package includes a Thesaurus to facilitate its usage by research teams across the globe, and working in different languages and with different recording traditions. The Thesaurus enables the *zoolog* package to recognises many different names for taxa and skeletal elements (e.g. ‘Bos taurus’, ‘Bos’, ‘BT’,  ‘bovino’, ‘bota’). Consequently, there is no need to use a particular, standardised recording code for the names of different taxa or elements. This first version of has benefited from the contributions from Moussab Besso, Canan Cakirlar, Jwana Chahoud, Jacopo De Grossi Mazzorin, Sabine Deschler-Erb, Dimitrios Filioglou, Armelle Gardeisen, Sierra Harding, Pilar Iborra, Michael MacKinnon, Nimrod Marom, Claudia Minniti, Francesca Slim, Barbara Stopp, and Emmanuelle Vila. We are grateful to them for their comments and help. In addition, users are encouraged to contribute to the Thesaurus so it can be expanded and adapted to any database.


## Installation

You can install the released version of zoolog from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("zoolog")
```

And the development version from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("josempozo/zoolog@HEAD")
```


# Reference

The log-ratios are calculated with respect to a standard reference. The package
**zoolog** includes three reference datasets. In addition, the user can consider other references, or the provided references can be extended and updated integrating newer research data. Submission of extended/improved references is encouraged. Please, contact the mantainer through the provided email address so the new reference is fully accessible and your name added to the list of contributors. 

The references are provided as dataframes named `ReferenceNietoAlbarellaDavis`, `ReferenceBasel`, and `ReferenceCombi`. They include
reference values for the main domesticates and red deer: *Bos taurus*, *Ovis aries*, *Capra hircus*, *Sus domesticus*, and *Cervus elaphus*:

* **ReferenceNietoDavisAlbarella** Includes the measurements from a female cow from Late Neolithic Minferri in Catalonia [@nieto2018element], the mean values of the adult female Shetland sheep measurements described in [@davis1996measurements], and the pig measurements from [@albarella2005neolithic].

* **ReferenceBasel** Includes the measurements compiled by Barbara Stopp from the reference collection in the Integrative Prähistorische und Naturwissenschaftliche Archäologie (IPNA, University of Basel, Switzerland). The specimens included are: *Bos taurus* Inv.nr. 2426 (Hinterwälder; female; 17 years old; live weight: 340 kg; withers height: 113 cm), *Ovis musimon* Inv.nr. 2266 (Male; adult), *Capra hircus* Inv.nr. 1597 (Male; adult), *Sus scrofa* Inv.nr. 1446 (male; 2-3 years old; life weight: 120 kg), *Cervus elaphus* Inv.nr. 2271 (male; adult). 

* **ReferenceCombi** Includes the most comprehensive reference for each species so more measurements can be considered: the Late Neolithic female cow from Minferri site in present-day Catalonia [@nieto2018element], the mean measurements from the Male Soay sheep (of known age) and mean measurements from male/female goats (age unknown) from [@clutton1990osteology], the pig measurements from Inv.nr. 1446 (male; 2-3 years old; life weight: 120 kg) from IPNA-Basel, and the red deer measurements from Inv.nr.2271 at IPNA-Basel registered by Barbara Stopp.    

The details of the reference data are described in \code{\link{References}}.

# Functions
The full list of functions is available under the 'Help' page of zoolog. Here we present the three main ones and describe their use:

* **`LogRatios`** This function computes the log ratios of the measurements relative to standard reference values. Three references are provided with the package. By default the 'ReferenceCombi' is used, as it is the most comprehensive so far. The function includes the option to 'joinCategories' allowing several taxa (typically Ovis, Capra, and unknown Ovis/Capra) to be considered together.

* **`RemoveNACases`** This function removes the table rows for which all measurements of interest are non-available (NA). A particular list of measurement names can be explicitly provided or selected by a common initial pattern (e.g. prefix). The default setting removes the rows with no available log ratios to facilitate subsequent analysis of the data.

* **`CondenseLogs`** This function condenses the calculated log ratio values into a reduced number of features by grouping log ratio values and selecting or calculating a feature value. By default the selected groups represent a single dimension, i.e. `Length` and `Width`. Only one feature is extracted per group. Currently, two methods are possible: "priority" (default) or "average". This operation is motivated by two circumstances. First, not all measurements are available for every bone specimen, which obstructs their direct comparison and statistical analysis. Second, several measurements can be strongly correlated (e.g. SD and Bd both represent bone width). Thus, considering them as independent would produce an over-representation of bone remains with multiple measurements per axis. Condensing each group of measurements into a single feature (e.g. one measure per axis) alleviates both problems. The default method ("priority"), selects the first available log ratio in each group. We provide the following by-default group and prioritization: For lengths, the first option is GL, then HTC. For widths, the order of priority is: Bd, BT, Bp, SD. This order maximises the robustness and reliability of the measurements, as priority is given to the most abundant, more replicable, and less age dependent measurements.The method 'average' extracts the mean per group, ignoring the non-available log ratios.

# Thesaurus

A thesaurus set is defined in order to make the methods robust to different nomenclatures used in datasets created by different authors and in different languages. The user can also use other thesaurus sets, or can modify the provided one. In this latter case, please, contact the maintainer at the provided email address so we can include the additions to the new versions of the package. Your name will be added to the list of contributors, as you will help to make the package more robust and comprehensive.

So far, four thesauri are included in **zoolog**:

* **`identifierThesaurus`** refers to the column names that will be considered to calculate the log ratios. So far, it includes the categories 'Taxon', 'Element', 'Measure', and 'Standard'. Each of these categories has different options (e.g.: 'Taxon' includes Taxon, TAX, species, animal, Specie, GenusSpecies, Especie, see below). This enables the `LogRatios` function to recognise the relevant column and compute the log ratios regardless of its name as long as it is included in the relevant Thesaurus (`taxonThesaurus`, `elementThesaurus`, `measureThesaurus`).

* **`taxonThesaurus`** refers to the various possibilities for naming taxa when recording animal bones. This enables the `LogRatios` function to recognise the name of each species and to find the relevant biometrical reference no matter what it is called in the database and the reference file (e.g. if in the user database, 'BT' refers to Bos taurus, but this is written 'bota' in the osteometrical reference). The `LogRatios` function correctly matches these terms because both categories ('BT' and 'bota') are included in the 'Bos taurus' category of the `taxonThesaurus`. To facilitate the comparison of names, the function has been made case, punctuation and spacing insensitive by default ('Bos' is considered same as 'bos' and 'Bos.').You can see the categories currently included as follows:

```{r}
library(zoolog)
thesaurusFile <- system.file("extdata", "taxonThesaurus.csv", package="zoolog")
thesaurusTax <- ReadThesaurus(thesaurusFile)
thesaurusTax
```

* **`elementThesaurus`** refers to the various possibilities for naming anatomical elements when recording animal bones. This enables the `LogRatios` function to recognise the name of each element and to find the relevant biometrical reference regardless of what it is called in the database and the reference file (e.g. if in the user database, 'omero' refers to humerus, but this is written 'HU' in the osteometrical reference). The `LogRatios` function correctly matches these terms because both categories ('omero' and 'HU') are included in the 'humerus' category of the `elementThesaurus`. To facilitate the comparison of names, the function has been made case, quotation and spacing insensitive by default ('Hum' is considered same as 'hum' and 'Hum.'). You can see the categories currently included as follows:

```{r}
thesaurusFile <- system.file("extdata", "elementThesaurus.csv", package="zoolog")
thesaurusEL <- ReadThesaurus(thesaurusFile)
thesaurusEL
```

* **`measureThesaurus`** refers to the various possibilities for naming measurements. In general, the English abbreviations from [@von1976guide] and [@davis1992rapid] are widely used. But this enables other nomenclatures (e.g. original German abbreviations in [@von1976guide]) to be included, or several measurements to be nested under the same category. In this case, GLl is considered as GL.    

# Examples
## Reading data and calculating log ratios
This example reads a dataset from a file in csv format and computes the
log-ratios. Then, the cases with no available log-ratios are removed.
Finally, the resulting dataset is saved in a file in csv format.

The first step is to set the local path to the folder where you have the dataset to be analysed (this is typically a comma-separated value (csv) file). Here the example dataset from [@valenzuela2008alimentacio] included in the package is used:

```{r}
library(zoolog)
dataFile <- system.file("extdata", "dataValenzuelaLamas2008.csv.gz", 
                        package = "zoolog")
data = read.csv2(dataFile,
                 quote="\"", header=TRUE, na="",
                 fileEncoding="UTF-8")
```

We now calculate the log-ratios using the function `LogRatios`. Only measurements that have an associated standard will be included in this calculation. The log values will appear as new columns with the prefix 'log' following the original columns with the raw measurements:
```{r}
dataWithLog=LogRatios(data)
head(dataWithLog)[,-c(6:20,32:63)]
```
We could be interested in obtaining the log ratios of all caprine, including *ovis aries*, *capra hircus*, and *ovis capra*, with respect to the reference for *ovis aries*. This can be set using the argument `joinCategories`.
```{r}
categories <- list(ovar = c("sheep", "capra", "oc"))
dataWithLog=LogRatios(data, joinCategories = categories)
head(dataWithLog)[,-c(6:20,32:63)]
```
Note that this option does not remove the distinction in the data between the different species, it just indicates that for these taxa the log ratios must be computed from the same reference (`ovar`).

The cases without log-ratios can be removed to facilitate subsequent analyses:
```{r}
dataWithLogPruned=RemoveNACases(dataWithLog)
```

You may want to write the resulting file in the working directory (you need to set it first):
```{r, eval = FALSE}
write.csv2(dataWithLogPruned, "myDataWithLogValues.csv", 
            quote=FALSE, row.names=FALSE, na="", 
            fileEncoding="UTF-8")
```

## Condensing log values
After calculating log ratios using the `LogRatios` function, many rows in the resultant dataframe (dataWithLog in the example above) may contain multiple log values, i.e. you will have several log values associated with a particular archaeological specimen. When analysing log ratios, it is preferential to avoid overrepresentation of bones with a greater number of measurements and account for each specimen only once. The `CondenseLogs` function extracts one length and one width value from each row and places these in new Length and Width columns.The 'priority' method described in [@trentacoste2018pre] has been set as default. Nevertheless, other options (e.g. average of all width log values for a given specimen) can be chosen if preferred. In this case, the default option has been used:
```{r}
dataWithSummary <- CondenseLogs(dataWithLogPruned)
head(dataWithSummary)[, -c(6:20,32:63)]
```



## Examples of different plots for data visualisation
Condensed log values can be visualised as histograms and box plots using ggplot [@wickham2011ggplot2] and tidyverse [@wickham2019welcome]. Here we’ll look at some examples of plotting values from sheep/goat.

### Preparing the data
The integration of the thesaurus functionality facilitates the use of datasets with heterogenous nomenclatures, without further preprocessing. An extensive catalogue of names for equivalent categories have been integrated in the provided thesaurus set `zoologThesaurus`. These equivalences are internally and silently managed without requiring any action from the user. However, it can be also interesting to explicitly standardize the data to make figures legible to a wider audience. This is especially useful when different nomenclature for the same concept is found in the same dataset, for instance "sheep" and "ovis" for the same taxon or "hum" and "HU" for the bone element.

In the example data, `zoologThesaurus` will change for instance `"ovar"` to `"ovis aries"`.
```{r}
dataStandardized <- StandardizeDataSet(dataWithSummary)
head(dataStandardized)[, -c(6:20,32:63)]
```

We may be interested in selecting all caprine elements. This can be done even without standardizing the data using the function `InCategory`:
```{r}
dataOC <- subset(dataWithSummary, InCategory(Especie, 
                                             c("sheep", "capra", "oc"),
                                             zoologThesaurus$taxon))
head(dataOC)[, -c(6:20,32:63)]
```
Observe that no standardization is performed in the output subset. To standardize the subset data, `StandardizeDataSet` can be applied either before or after the subsetting.
```{r}
dataOCStandardized <- StandardizeDataSet(dataOC)
head(dataOCStandardized)[, -c(6:20,32:63)]
```
Observe also that the distinction between "ovis aries", "capra hircus", and "ovis capra" has not been removed from the data.

For plotting the Width, we can retain the cases including this measure:
```{r}
dataOCWithWidth <- RemoveNACases(dataOCStandardized, measureNames = "Width")
```
 which gives
```{r}
nrow(dataOCWithWidth)
```
cases.

For ploting the Length, we can retain the cases including this measure:
```{r}
dataOCWithLength <- RemoveNACases(dataOCStandardized, measureNames = "Length")
```

which gives 
```{r}
nrow(dataOCWithLength)
```
cases.

### Plotting the data

#### Horizontal Boxplot with dots
For the example plots we will use the package **ggplot2**.
```{r, echo = FALSE}
library(ggplot2)
```

We can now create a boxplot for the widths: 
```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
ggplot(dataOCStandardized, aes(x=Width, y=Site)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height=.2, width=0, alpha=1/2, color=4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio")
```

And another boxplot for the lengths:
```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
ggplot(dataOCStandardized, aes(x=Length, y=Site)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height=.2, width=0, alpha=1/2, color=4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine lengths") +
  xlab("Length log-ratio")
```

#### Histograms 
We may choose to plot the width data as an histogram: 
```{r, fig.asp = 0.7, fig.width = 6, fig.align="center"}
ggplot(dataOCStandardized, aes(Width)) +
  geom_histogram(bins=30, na.rm = TRUE) +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio") +
  facet_grid(Site ~.) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  theme(plot.title = element_text(hjust=0.5,size=14),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text = element_text(size=10) ) +
  scale_y_continuous(breaks=c(0,10,20,30))
```

We may run a statistical test (here a Student t-test) to check whether the differences between sites have statistical significance on widths:
```{r}
t.test(Width ~ Site, dataOCStandardized, 
       subset = Site %in% c("OLD", "ALP"))
```
and/ or in lengths:
```{r}
t.test(Length ~ Site, dataOCStandardized, 
       subset = Site %in% c("OLD", "ALP"))
```

#### Another visualisation for boxplot with dots 
Here we reorder the factor levels of `dataOCStandardized$Taxon` to make
the order of the boxplots more intuitive.

```{r}
levels0 <- levels(dataOCStandardized$Taxon)
levels0
```
```{r}
dataOCStandardized$Taxon <- factor(dataOCStandardized$Taxon, 
                                   levels = levels0[c(2,3,1)])
levels(dataOCStandardized$Taxon)
```
and assign specific colours for each category:
```{r, warning = FALSE, fig.asp = 0.6, fig.width = 6, fig.align="center"}
Ocolour <- c("#A2A475", "#D8B70A", "#81A88D")
ggplot(dataOCStandardized, aes(x=Site, y=Width)) + 
  geom_boxplot(aes(fill=Taxon), 
               notch=TRUE, alpha=0, lwd=0.377, outlier.alpha = 0,
               width = 0.5,
               position=position_dodge(0.75), na.rm = TRUE,
               show.legend = FALSE) + 
  geom_point(aes(colour=Taxon, shape=Taxon), 
             alpha=0.7, size=0.8, 
             position = position_jitterdodge(jitter.width=.3),
             na.rm = TRUE) +
  scale_colour_manual(values=Ocolour) +
  scale_shape_manual(values=c(15, 18, 16)) +
  theme_bw(base_size = 8) +
  ylab("LSI value") +
  ggtitle("Sheep/goat LSI width values") 
```


#### Another visualisation for sheep and goat histograms:  

```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
TaxonSiteWidthHist <- ggplot(dataOCStandardized, aes(Width, fill=Taxon)) + 
  geom_histogram(bins=30, alpha=.5, position="identity") + 
  ggtitle("Sheep/goat Widths") + facet_grid(Site ~ Taxon)
TaxonSiteWidthHist
```
```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
TaxonSiteWidthHist <- ggplot(dataOCStandardized, aes(Width, fill=Taxon)) + 
  geom_histogram(bins=30, alpha=.5, position="identity") + 
  ggtitle("Sheep/goat Widths") + facet_grid(~Site)
TaxonSiteWidthHist
```




