---
title: "Package ***zoolog***:  \n zooarcheological analysis with log-ratios."
author: "Jose M Pozo, Silvia Valenzuela-Lamas, Angela Trentacoste,  Ariadna Nieto-Espinet and Silvia Guimar찾es Chiarelli"
email: "svalenzuela@imf.csic.es"
date: "`r format(Sys.Date())`"
bibliography: zoolog.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zoolog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The package ***zoolog*** includes functions and reference data to generate and 
handle log-ratios from measurements obtained on zooarchaeological material
following the seminal works from [@von1976guide] and [@davis1992rapid]. The 
package includes several sets of standard reference values.

The methods included in the package were first developed in the frame of the 
ERC-Starting Grant 716298 ZooMWest (PI S. Valenzuela-Lamas), and were first used 
in the paper [@trentacoste2018pre].

In general, zooarcheological datasets are composed by bone remains of many  different anatomical body parts. The analysis of measurements from a given  anatomical element provides the best control for the variables affecting size and shape and, as such, it is the preferable option. Unfortunately, this approach is  not always viable due to low sample sizes in some archaeological assemblages. Length and width measurements of different anatomical elements cannot be directly compared or aggregated for statistical analysis. This problem can be palliated by extracting the measurement log-ratios with respect to a reference. The resulting log-ratios can be compared and statistically analysed under general reasonable  conditions [@albarella2002size].

The package also includes a Thesaurus to facilitate its usage by research teams across the globe and working in different languages and with different recording traditions. Consequently, there is no need to use a particular recording code for *zoolog* for it to recognise the names of the species or the anatomical element.  This first version has benefited from the contributions from Moussab Besso, Canan Cakirlar, Jwana Chahoud, Jacopo De Grossi, Dimitrios Filioglou, Armelle Gardeisen, Sierra Harding, Pilar Iborra, Michael MacKinnon, Nimrod Marom, Claudia Minniti, Francesca Slim, Emmanuelle Vila. We are grateful to them for their comments and help. In addition, users are encouraged to contribute to the Thesaurus so it can be expanded and adapted to any database.


## Installation

You can install the released version of zoolog from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("zoolog")
```

And the development version from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("josempozo/zoolog")
```


# Reference

The log-ratios are calculated with respect to a standard reference. The package
*zoolog* includes three reference datasets. In addition, the user can consider other references, or the provided references can be extended and updated integrating newer research data. Submission of extended/improved references is encouraged. Please, contact the mantainer through the provided email address so the new reference is fully accessible and your name added to the list of contributors. 

The references are provided as dataframe named `ReferenceNietoAlbarellaDavis`, `ReferenceBasel`, and `ReferenceCombi`. They include
reference values for the main domesticates and reed deer: *Bos taurus*, *Ovis aries*, *Capra hircus*, *Sus domesticus*, and *Cervus elaphus*:

**ReferenceNietoDavisAlbarella** Includes the measurements from a female cow from Late Neolithic Minferri in Catalonia [@nieto2018element], the mean values of the adult female Shetland sheep measurements described in [@davis1996measurements], and the pig measurements from [@albarella2005neolithic].

**ReferenceBasel** Includes the measurements compiled by Barbara Stopp from the reference collection in the Integrative Pr채historische und Naturwissenschaftliche Arch채ologie (IPNA, University of Basel, Switzerland). The specimens included are: *Bos taurus* Inv.nr. 2426 (Hinterw채lder; female; 17 years old; live weight: 340 kg; withers height: 113 cm), *Ovis musimon* Inv.nr. 2266 (Male; adult), *Capra hircus* Inv.nr. 1597 (Male; adult), *Sus scrofa* Inv.nr. 1446 (male; 2-3 years old; life weight: 120 kg), *Cervus elaphus* Inv.nr. 2271 (male; adult). 

**ReferenceCombi** Includes the measurements from the Late Neolithic female cow from Minferri site in present-day Catalonia [@nieto2018element], the mean measurements from the Soay sheep males aged and mean measurements from male/female not aged goats from [@clutton1990osteology], the pig measurements from [@albarella2005neolithic], and the red deer measurements from Inv.nr.2271 at IPNA-Basel registered by Barbara Stopp.   

The details of the reference data format are described in \code{\link{References}}.

# Functions

# Thesaurus

# Examples

This example reads a dataset from a file in csv format and compute the
log-ratios. Then, the cases with no available log-ratios are removed.
Finally, the resulting dataset is saved in a file in csv format.

The first step is to set the local path where you have the dataset to be analysed (this is typically a comma-separated value (csv) file). Here the example dataset from [@valenzuela2008alimentacio] included in the package is used:

```{r}
library(zoolog)
dataFile <- system.file("extdata", "dataValenzuelaLamas2008.csv.gz", 
                        package = "zoolog")
data = read.csv2(dataFile,
                 quote="\"", header=TRUE, na="",
                 fileEncoding="UTF-8")
```

We now calculate the log-ratios using the function LogRatios(). The log values will appear as new columns next to the original ones with raw measurements:
```{r}
dataWithLog=LogRatios(data)
head(dataWithLog)[,-c(6:20,32:63)]
```

The cases without log-ratios can be removed to ease the later analyses:
```{r}
dataWithLogPruned=RemoveNACases(dataWithLog)
```

You may want to write the resulting file in the working directory (you need to set it before):
```{r}
write.csv2(dataWithLogPruned, "myDataWithLogValues.csv", 
            quote=FALSE, row.names=FALSE, na="", 
            fileEncoding="UTF-8")
```



## Examples of different plots for data visualisation
library(zoolog)
library("tidyverse")

The function CondenseLogs () condenses the measurements per axis (e.g. 'lenghts', 'widths', 'depths'). The 'priority' method described in [@trentacoste2018pre] has been set as default. Nevertheless, other options (e.g. average of all width log values for a given specimen) can be chosen if preferred. In this case, the default option has been used:
```{r}
dataWithSummary <- CondenseLogs(dataWithLogPruned)
head(dataWithSummary)[, -c(6:20,32:63)]
```

The function zoologThesaurus standardizes the data, so it is possible to consider different names for a given taxon and element (e.g. sheep and ovis; hum and HU) present in the dataframe wo there is no need of homogenizing the dataframe before the analysis:
```{r}
dataStandardized <- StandardizeDataSet(dataWithSummary)
## This allows to select the caprine cases by
dataOC <- subset(dataStandardized, Taxon=="ovis aries")

## For ploting the Width, we can retain the cases including this measure:
dataOCWithWidth <- RemoveNACases(dataOC, measureNames = "Width")
## including
nrow(dataOCWithWidth)
## cases.
## For ploting the Width, we can retain the cases including this measure:
dataOCWithLength <- RemoveNACases(dataOC, measureNames = "Length")
## including
nrow(dataOCWithLength)
## cases.

## Outlier in dataOC[27,] (logSD = 0.4649724)

########### Horizontal Boxplot with dots
library("ggplot2")
outFolder <- "~/Silvia/Figures"

ggplot(dataOC, aes(x=Width, y=Jaciment)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height=.2, width=0, alpha=1/2, color=4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio")

ggplot(dataOC, aes(x=Length, y=Jaciment)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height=.2, width=0, alpha=1/2, color=4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine lengths") +
  xlab("Length log-ratio")


########### Histograms Caprines Widths

ggplot(dataOC, aes(Width)) +
  geom_histogram(bins=30, na.rm = TRUE) +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio") +
  facet_grid(Jaciment ~.) +
  theme_bw() +
  theme(	panel.grid.major.y = element_blank(),
         panel.grid.minor.y = element_blank()) +
  theme(plot.title = element_text(hjust=0.5,size=22),
        axis.title.x = element_text(size=18),
        axis.title.y = element_text(size=18),
        axis.text = element_text(size=12) ) +
  scale_y_continuous(breaks=c(0,10,20,30))


ggplot(dataOC, aes(Length)) +
  geom_histogram(bins=30, na.rm = TRUE) +
  ggtitle("Caprine lengths") +
  xlab("Length log-ratio") +
  facet_grid(Jaciment ~.) +
  theme_bw() +
  theme(	panel.grid.major.y = element_blank(),
         panel.grid.minor.y = element_blank()) +
  theme(plot.title = element_text(hjust=0.5,size=22),
        axis.title.x = element_text(size=18),
        axis.title.y = element_text(size=18),
        axis.text = element_text(size=12) ) +
  scale_y_continuous(breaks=c(0,10,20,30))


t.test(Width ~ Jaciment, dataOC, subset = Jaciment %in% c("TFC", "ALP"))

####################################################
```



```{r ibahernando, out.width="60%", fig.width=8, fig.asp=750/666, fig.align="center", warning=FALSE, fig.cap="Ibahernando stelae (decoration 5) with only *normal* edges, node 1  overlaps node 2 and node 3"}
ggplot(dataOC, aes(x=Width, y=Jaciment)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height=.2, width=0, alpha=1/2, color=4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio")
```


## Sheep.goat jitter and box plots (from AAS article) ------------------

# From dataframe with all logs (Ndata)
#DOTPLOT
# from data select TAX sheep (OVA), goat (CAH), and sheep/goat (O)
# graph period vs width values
# vary point colour and shape with TAX
# set point colour to Ocolour
# set shapes manually
# if points are to close together or far apart change 'jitter.width=.3'
# BOXPLOT
# from Ndata only graph sheep (OVA) and goat (CAH)
# outlier.alpha = 0 so ggplot will not plot black dot for outliers


Ocolour <- c("#A2A475", "#D8B70A", "#81A88D")

OC_W_Jitter <- ggplot(data=subset(Ndata, TAX=="OVA" | TAX=="O" | TAX=="CAH"), 
                      aes(x=Period, y=Width)) + 
  geom_point(data=subset(Ndata, TAX=="OVA" | TAX=="O" | TAX=="CAH"), 
             aes(colour=TAX, shape=TAX), alpha=0.7, size=0.8, position = position_jitterdodge(jitter.width=.3)) +
  scale_colour_manual(values=Ocolour) +
  scale_shape_manual(values=c(15, 18, 16)) +
  geom_boxplot(data=subset(Ndata, TAX=="OVA" | TAX=="CAH"), aes(x=Period, y=Width, fill=TAX), 
               notch=TRUE, alpha=0,  lwd=0.377, outlier.alpha = 0, position=position_dodge(1)) + 
  theme_bw(base_size = 8) +
  ylab("LSI value")+
  theme(axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        plot.title = element_text(size=8)) +
  ggtitle("Sheep/goat LSI width values") 
OC_W_Jitter



## Sheep.goat histogram (from PLOS) -----------------


# subset dataframe (here called ItalyLogComplemented) to only O, OVA, and CAH in ItalyLogComplemented$TAX
# make histogram and facet by region and period (period here is called Chron2)
# alpha=.5 makes the histograms transparent, colours are default

SGLengthHist <- ggplot(
  subset(ItalyLogComplemented, TAX=="OVA" | TAX=="O" | TAX=="CAH" ), 
  aes(Length, fill=TAX)) + 
  geom_histogram(bins=50,alpha=.5, position="identity") + 
  ggtitle("Sheep/goat Lengths") + facet_grid(Chron2~Region)


#To add mean line, calculate log means for each period (here Chron2)
#columns 23 and 24 in ItalyLogComplemented = ItalyLogComplemented$Length and ItalyLogComplemented$Width (the selected length and eidth logs for each row)
# >>>there is probably an easier way to do this in tidyverse using summarise()?
LogMeansChron2 <- aggregate(ItalyLogComplemented[,23:24], 
                            list(Region = ItalyLogComplemented$Region, 
                                 Chron2 = ItalyLogComplemented$Chron2, 
                                 TAX = ItalyLogComplemented$TAX), 
                                FUN=mean, na.rm=TRUE)

#Add mean line and label to each facet
# geom_vline adds a vertical line 
# only for TAX == OVA and CAH (not interderminate sheep/goat)
# line type and colour vary with TAX
# line is placed at mean (xintercept=LogMeansChron2$Length)
# geom_text adds label with mean only for OVA
# label is rounded to 3 decimal places
# label is placed on x-axis at mean value + 0.025, y-axis at 14
SGLengthHist + geom_vline(data=subset(LogMeansChron2, TAX=="OVA" | TAX=="CAH"), 
                          aes(xintercept=Length, linetype=TAX, colour=TAX) , size=0.25) + 
  scale_linetype_manual(values=c("longdash", "solid")) + 
  geom_text(data=subset(LogMeansChron2, TAX=="OVA"), 
            aes(label = round(Length, digits=3), y=14, x=Length+0.025), size = 3)




