---
title: "Package ***zoolog***:  \n zooarcheological analysis with log-ratios."
author: "Jose M Pozo, Silvia Valenzuela-Lamas, Angela Trentacoste,  Ariadna Nieto-Espinet and Silvia Guimarães Chiarelli"
email: "svalenzuela@imf.csic.es"
date: "`r format(Sys.Date())`"
bibliography: zoolog.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zoolog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The package ***zoolog*** includes functions and reference data to generate and 
manipulate log-ratios (also known as log size index (LSI) values) from measurements obtained on zooarchaeological material. Log ratios are used to compare the relative (rather than the absolute) dimensions of animals from archaeological contexts [@meadow1999use]. Essentially, the method compares archaeological measurements to a standard, producing a value that indicates how much larger or smaller the archaeological specimen is compared to that standard.

The methods included in the package were first developed in the framework of the 
ERC-Starting Grant 716298 ZooMWest (PI S. Valenzuela-Lamas), and were first used 
in the paper [@trentacoste2018pre]. They are based on the techniques proposed by [@simpson1941large] and [@simpson1960quantitative], which calculates LSI values as such: 
LSI (log size index) = (log x - log m) = log(x/m)


The zoolog package is based on the measurements and measurement abbreviations defined in [@von1976guide] and [@davis1992rapid]. Several different sets of standard reference values for calculating the log ratios are included in the package. These standards included several published and widely used biometric datasets (i.e. [@davis1996measurements], [@albarella2005neolithic]) as well as other less known. We are particularly grateful to Sabine Deschler-Erb and Barbara Stopp, from the University of Basel (Switzerland) for making the reference values of several specimens available through the ICAZ Roman Period Working Group, which have been included here with their permission. 

In general, zooarcheological datasets are composed by skeletal remains representing many different anatomical body parts. In investigation of animal size, the analysis of measurements from a given anatomical element provides the best control for the variables affecting size and shape and, as such, it is the preferable option. Unfortunately, this approach is not always viable due to low sample sizes in some archaeological assemblages. This problem can be mitigated by calculating the LSI values for measurements with respect to a reference, which provides a means of aggregating biometric information from different body parts. The resulting log ratios can be compared and statistically analysed under general reasonable conditions [@albarella2002size]. However, length and width measurements of different anatomical elements still should not be directly compared or aggregated for statistical analysis. 

The package includes a Thesaurus to facilitate its usage by research teams across the globe, and working in different languages and with different recording traditions. The Thesaurus enables the *zoolog* package to recognises many different names for taxa and skeletal elements (e.g. ‘Bos taurus’, ‘Bos’, ‘BT’,  ‘bovino’, ‘bota’). Consequently, there is no need to use a particular, standardised recording code for the names of different taxa or elements. This first version of has benefited from the contributions from Moussab Besso, Canan Cakirlar, Jwana Chahoud, Jacopo De Grossi Mazzorin, Sabine Deschler-Erb, Dimitrios Filioglou, Armelle Gardeisen, Sierra Harding, Pilar Iborra, Michael MacKinnon, Nimrod Marom, Claudia Minniti, Francesca Slim, Barbara Stopp, and Emmanuelle Vila. We are grateful to them for their comments and help. In addition, users are encouraged to contribute to the Thesaurus so it can be expanded and adapted to any database.


## Installation

You can install the released version of zoolog from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("zoolog")
```

And the development version from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("josempozo/zoolog")
```


# Reference

The log-ratios are calculated with respect to a standard reference. The package
*zoolog* includes three reference datasets. In addition, the user can consider other references, or the provided references can be extended and updated integrating newer research data. Submission of extended/improved references is encouraged. Please, contact the mantainer through the provided email address so the new reference is fully accessible and your name added to the list of contributors. 

The references are provided as dataframes named `ReferenceNietoAlbarellaDavis`, `ReferenceBasel`, and `ReferenceCombi`. They include
reference values for the main domesticates and reed deer: *Bos taurus*, *Ovis aries*, *Capra hircus*, *Sus domesticus*, and *Cervus elaphus*:

* **ReferenceNietoDavisAlbarella** Includes the measurements from a female cow from Late Neolithic Minferri in Catalonia [@nieto2018element], the mean values of the adult female Shetland sheep measurements described in [@davis1996measurements], and the pig measurements from [@albarella2005neolithic].

* **ReferenceBasel** Includes the measurements compiled by Barbara Stopp from the reference collection in the Integrative Prähistorische und Naturwissenschaftliche Archäologie (IPNA, University of Basel, Switzerland). The specimens included are: *Bos taurus* Inv.nr. 2426 (Hinterwälder; female; 17 years old; live weight: 340 kg; withers height: 113 cm), *Ovis musimon* Inv.nr. 2266 (Male; adult), *Capra hircus* Inv.nr. 1597 (Male; adult), *Sus scrofa* Inv.nr. 1446 (male; 2-3 years old; life weight: 120 kg), *Cervus elaphus* Inv.nr. 2271 (male; adult). 

* **ReferenceCombi** Includes the measurements from the Late Neolithic female cow from Minferri site in present-day Catalonia [@nieto2018element], the mean measurements from the Soay sheep males aged and mean measurements from male/female not aged goats from [@clutton1990osteology], the pig measurements from [@albarella2005neolithic], and the red deer measurements from Inv.nr.2271 at IPNA-Basel registered by Barbara Stopp.   

The details of the reference data format are described in \code{\link{References}}.

# Functions

# Thesaurus

# Examples
## Reading data and calculating log ratios
This example reads a dataset from a file in csv format and computes the
log-ratios. Then, the cases with no available log-ratios are removed.
Finally, the resulting dataset is saved in a file in csv format.

The first step is to set the local path to the folder where you have the dataset to be analysed (this is typically a comma-separated value (csv) file). Here the example dataset from [@valenzuela2008alimentacio] included in the package is used:

```{r}
library(zoolog)
dataFile <- system.file("extdata", "dataValenzuelaLamas2008.csv.gz", 
                        package = "zoolog")
data = read.csv2(dataFile,
                 quote="\"", header=TRUE, na="",
                 fileEncoding="UTF-8")
```

We now calculate the log-ratios using the function LogRatios(). Only measurements that have an associated standard will be included in this calculation. The log values will appear as new columns following the original ones with raw measurements:
```{r}
dataWithLog=LogRatios(data)
head(dataWithLog)[,-c(6:20,32:63)]
```

The cases without log-ratios can be removed to ease the later analyses:
```{r}
dataWithLogPruned=RemoveNACases(dataWithLog)
```

You may want to write the resulting file in the working directory (you need to set it before):
```{r, eval = FALSE}
write.csv2(dataWithLogPruned, "myDataWithLogValues.csv", 
            quote=FALSE, row.names=FALSE, na="", 
            fileEncoding="UTF-8")
```

## Condensing log values
After calculating log ratios using the LogRatios() function, many rows in the resultant dataframe (dataWithLog in the example above) may contain multiple log values, i.e. you will have several log values associated with a particular archaeological specimen. When analysing log ratios, it is preferential to avoid overrepresentation of bones with a greater number of measurements and account for each specimen only once. The CondenseLogs() extracts one length and one width value from each row and places these in new Length and Width columns.The 'priority' method described in [@trentacoste2018pre] has been set as default. Nevertheless, other options (e.g. average of all width log values for a given specimen) can be chosen if preferred. In this case, the default option has been used:
```{r}
dataWithSummary <- CondenseLogs(dataWithLogPruned)
head(dataWithSummary)[, -c(6:20,32:63)]
```



## Examples of different plots for data visualisation
Condensed log values can be visualised as histograms and box plots using ggplot [@wickham2011ggplot2] and tidyverse [@wickham2019welcome]. Here we’ll look at some examples of plotting values from sheep/goat.

### Preparing the data
???To make the figure legible to a wider audience, we can standardize the data according to the *zoologThesaurus*. 
???The function zoologThesaurus standardizes the data, so it is possible to consider different names for a given taxon and element (e.g. sheep and ovis; hum and HU) present in the dataframe

In the example data, this will change the code for sheep/goat from ‘oc’ to ‘ovis aries’
```{r}
dataStandardized <- StandardizeDataSet(dataWithSummary)
```

This allows to select the caprine cases by
```{r}
dataOC <- subset(dataStandardized, Taxon=="ovis aries")
```

For plotting the Width, we can retain the cases including this measure:
```{r}
dataOCWithWidth <- RemoveNACases(dataOC, measureNames = "Width")
```
 which gives
```{r}
nrow(dataOCWithWidth)
```
cases.

For ploting the Length, we can retain the cases including this measure:
```{r}
dataOCWithLength <- RemoveNACases(dataOC, measureNames = "Length")
```

which gives 
```{r}
nrow(dataOCWithLength)
```
cases.

### Plotting the data

#### Horizontal Boxplot with dots
It is first necessary to load the relevant libraries (ggplot2 and tidyverse). It is also advised to create a folder for the outputs:
```{r}
library (ggplot2)
library(tidyverse)
```

We can now create a boxplot for the widths: 
```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
ggplot(dataOC, aes(x=Width, y=Jaciment)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height=.2, width=0, alpha=1/2, color=4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio")
```

And another boxplot for the lengths:
```{r, fig.asp = 0.6, fig.width = 6, fig.align="center"}
ggplot(dataOC, aes(x=Length, y=Jaciment)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE) +
  geom_jitter(height=.2, width=0, alpha=1/2, color=4, na.rm = TRUE) +
  theme_bw() +
  ggtitle("Caprine lengths") +
  xlab("Length log-ratio")
```

#### Histograms 
We may choose to plot the width data as an histogram: 
```{r, fig.asp = 0.7, fig.width = 6, fig.align="center"}
ggplot(dataOC, aes(Width)) +
  geom_histogram(bins=30, na.rm = TRUE) +
  ggtitle("Caprine widths") +
  xlab("Width log-ratio") +
  facet_grid(Jaciment ~.) +
  theme_bw() +
  theme(	panel.grid.major.y = element_blank(),
         panel.grid.minor.y = element_blank()) +
  theme(plot.title = element_text(hjust=0.5,size=14),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text = element_text(size=10) ) +
  scale_y_continuous(breaks=c(0,10,20,30))
```

We may run a statistical test (here a Student t-test) to check whether the differences between sites have statistical significance:
```{r}
t.test(Width ~ Jaciment, dataOC, subset = Jaciment %in% c("TFC", "ALP"))
```

## Example plots from Angela


Sheep.goat jitter and box plots (from AAS article) ------------------

From dataframe with all logs (Ndata)

### DOTPLOT 
* from data select TAX sheep (OVA), goat (CAH), and sheep/goat (O)
* graph period vs width values
* vary point colour and shape with TAX
* set point colour to Ocolour
* set shapes manually
* if points are to close together or far apart change 'jitter.width=.3'

### BOXPLOT
* from Ndata only graph sheep (OVA) and goat (CAH)
* outlier.alpha = 0 so ggplot will not plot black dot for outliers

```{r}
Ocolour <- c("#A2A475", "#D8B70A", "#81A88D")

OC_W_Jitter <- ggplot(data=subset(Ndata, TAX=="OVA" | TAX=="O" | TAX=="CAH"), 
                      aes(x=Period, y=Width)) + 
  geom_point(data=subset(Ndata, TAX=="OVA" | TAX=="O" | TAX=="CAH"), 
             aes(colour=TAX, shape=TAX), alpha=0.7, size=0.8, position = position_jitterdodge(jitter.width=.3)) +
  scale_colour_manual(values=Ocolour) +
  scale_shape_manual(values=c(15, 18, 16)) +
  geom_boxplot(data=subset(Ndata, TAX=="OVA" | TAX=="CAH"), aes(x=Period, y=Width, fill=TAX), 
               notch=TRUE, alpha=0,  lwd=0.377, outlier.alpha = 0, position=position_dodge(1)) + 
  theme_bw(base_size = 8) +
  ylab("LSI value")+
  theme(axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        plot.title = element_text(size=8)) +
  ggtitle("Sheep/goat LSI width values") 
OC_W_Jitter
```


## Sheep.goat histogram (from PLOS) -----------------

* Subset dataframe (here called ItalyLogComplemented) to only O, OVA, and CAH in ItalyLogComplemented$TAX
* make histogram and facet by region and period (period here is called Chron2)
* alpha=.5 makes the histograms transparent, colours are default

```{r}
SGLengthHist <- ggplot(
  subset(ItalyLogComplemented, TAX=="OVA" | TAX=="O" | TAX=="CAH" ), 
  aes(Length, fill=TAX)) + 
  geom_histogram(bins=50,alpha=.5, position="identity") + 
  ggtitle("Sheep/goat Lengths") + facet_grid(Chron2~Region)
```

To add mean line, calculate log means for each period (here Chron2)
columns 23 and 24 in ItalyLogComplemented = ItalyLogComplemented$Length and ItalyLogComplemented$Width (the selected length and eidth logs for each row)
# >>>there is probably an easier way to do this in tidyverse using summarise()?
LogMeansChron2 <- aggregate(ItalyLogComplemented[,23:24], 
                            list(Region = ItalyLogComplemented$Region, 
                                 Chron2 = ItalyLogComplemented$Chron2, 
                                 TAX = ItalyLogComplemented$TAX), 
                                FUN=mean, na.rm=TRUE)

#Add mean line and label to each facet
# geom_vline adds a vertical line 
# only for TAX == OVA and CAH (not interderminate sheep/goat)
# line type and colour vary with TAX
# line is placed at mean (xintercept=LogMeansChron2$Length)
# geom_text adds label with mean only for OVA
# label is rounded to 3 decimal places
# label is placed on x-axis at mean value + 0.025, y-axis at 14
SGLengthHist + geom_vline(data=subset(LogMeansChron2, TAX=="OVA" | TAX=="CAH"), 
                          aes(xintercept=Length, linetype=TAX, colour=TAX) , size=0.25) + 
  scale_linetype_manual(values=c("longdash", "solid")) + 
  geom_text(data=subset(LogMeansChron2, TAX=="OVA"), 
            aes(label = round(Length, digits=3), y=14, x=Length+0.025), size = 3)




